---
title: "Causation & Randomized Experiments"
author: "Lachlan Deer"
institute: "Social Media and Web Analytics, Spring 2024"
format: 
  beamer: 
    aspectratio: 32
    navigation: horizontal
    theme: cousteau
---

```{r, echo=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(infer)
library(dagitty)
library(ggdag)
```

## Learning Goals

* Explain the challenges of causal inference with observational data 
* Define the term randomized experiment
* Explain the fundamental problem of causal inference
* Describe the Potential Outcomes framework for casual inference
* Define the Sample Average Treatment Effect
* Analyze data from a randomized experiment to estimate the sample average treatment effect of an intervention using statistical inference and linear regression

# Causation 

## Causal Questions

**Does X cause Y?** 
Examples of causal questions include:

* Does smoking cause cancer?
* Does exercise make people happier?
* Does my social media advertising increase sales?
* Does hiring an influencer to promote a product lead to an increased consumer awareness about the product?

Not all causal questions use the word "*cause*". Other words that imply causality include:

* Improve
* Increase / decrease
* Lead to
* Make

Tell-tale sign that a question is causal: analysis is used to make an argument for changing a procedure, policy, or practice.

## Association vs Causation

\begin{figure}
\centering
\includegraphics[width=12cm]{figs/nicholas_cage_drowning.png}
\end{figure}

## The Difficulty of Casual Effects in Observational Data

**Challenge 1**: Omitted Variables (Z): variable that affects both X & Y that is not included in the analysis

```{r, echo = FALSE, cache = TRUE, out.width="60%", fig.align='center'}

dag <- dagify(y ~ x + z,
  x ~ z,
  exposure = "x",
  outcome = "y"
)

ggdag(dag, 
      layout = "circle",
      node_size = 20,
      text_size = 10) +
    geom_dag_edges(edge_width = 2, 
                   #alpha = 0.4,
                   arrow_directed = grid::arrow(length = grid::unit(14, "pt"), type = "closed"),
                   arrow_bidirected = grid::arrow(length = grid::unit(14, "pt"), 
                                                  ends = "both", type ="closed"
                                                  ),
    
                   ) + 
    theme_dag_blank()


```

## Omitted Variable Bias

:::: {.columns}

::: {.column width="50%"}
\begin{figure}
\includegraphics[width=6cm]{figs/ovb.png}
\end{figure}
:::

::: {.column width="50%"}
\vspace{2cm}

Does exercise cause weightloss?
:::

::::

## The Difficulty of Casual Effects in Observational Data

**Challenge 2**: Selection Effects: improper (non random) selection of individuals, such that the sample of cases and controls are not drawn from the same reference population

```{r, cache = TRUE, out.width="60%", fig.align='center'}
dag <- dagify(y ~ x,
              y ~~ z,
  x ~~ z,
  exposure = "x",
  outcome = "y"
)

ggdag(dag, 
      layout = "circle",
      node_size = 20,
      text_size = 10) +
    geom_dag_edges(edge_width = 2, 
                   #alpha = 0.4,
                   arrow_directed = grid::arrow(length = grid::unit(14, "pt"), type = "closed"),
                   arrow_bidirected = grid::arrow(length = grid::unit(14, "pt"), 
                                                  ends = "both", type ="closed"
                                                  ),
    
                   ) + 
    theme_dag_blank()

```

## Selection Effects

:::: {.columns}

::: {.column width="50%"}
\begin{figure}
\includegraphics[width=6cm]{figs/selection.png}
\end{figure}
:::

::: {.column width="50%"}
\vspace{1cm}

Should we use to plan where to put additional armor on fighter jets based on the damage of planes that return to base? 
:::

::::


## Solutions to OVB and Selection Effects

Omitted Variable Bias and Selection Effects mean **our estimates** of the effect of X on Y **are biased**

Possible Solutions:

1. Modelling and/or assumptions 
2. Randomization of the intervention

# Randomization

## What is Randomization?

When we use the word **random** in this context, we mean:

* Every unit has some chance (i.e., a non-zero probability) of being selected to receive the intervention or be in the control group.
* The selection into these groups is based upon a random process

## The "Magic" of Randomization

Because our unit of analyses are **randomized to treatment and control groups**, **on average** there is **no difference** between these two groups on any **characteristics** other than their treatment.

Prior to treatment, on average the groups (Treatment and Control) are equivalent to one another on every observed and unobserved variable

* There is no omitted variable bias 
* There are no selection effects

Can we check for randomization?

# Application: Encouraging Donors to Share About Charity

##

\begin{figure}
\includegraphics[width=12cm]{figs/pymwymi.png}
\end{figure}

\footnotesize

Discussion based on: Silver & Small. (2023). [Put Your Mouth Where Your Money Is: A Field Experiment Encouraging Donors to Share About Charity](https://doi.org/10.1287/mksc.2023.1450)

## Intervention Context

* \textbf{\alert{Research Question}}: Can we effectively get donors to share about charitable donations?

* \textbf{\alert{Why is this relevant?}}
    * Raises awareness and bolsters fundraising efforts for good causes

* This isn't easy: Donors face a trade off
    - -ve: (Personal) Reputation risks via appearing braggy or inauthentic, vs.
    - +ve: (External) Word of Mouth benefits to the charity

* Today we'll explore this question in the context of charitable giving for educational projects, and an intervention that encourages sharing about the charity after donation
    * $Y_i =$ clicking on a sharing pop-up OR recruiting future donors (0/1)
    * $T_i =$ an intervention encouraging sharing about cause post-donation (0/1)

## Importing the Data

```{r, message=FALSE, echo =TRUE}
charity <- 
    read_csv("data/exp2data.csv") %>%
    mutate(condition = if_else(condition ==1, 
                               "treatment", 
                               "control"
                               )
    )
```

## Inspecting the Data
\footnotesize

```{r, echo = TRUE}
head(charity, n = 5)
```

## What is our data?

**Units**: objects being studied.

* Usually the rows of the data set.
* Examples: Survey respondents, consumers, firms, app users, influencers.
* Today's data: Charity Donors from DonorsChoose.org

**Variables**: measurements that can vary across units.

* Usually the columns of a data set.
* Examples: quantity bought, dollars spent, participation in an experiment.
* Today's data: Amount donated, clicked through pop up, referred a friend, treatment status, 

# Estimating the Treatment Effect

## Causal Questions 

\alert{Does} cause based solicitation *increase* sharing? 

\alert{Does} cause based solicitation recruit *more* future donors?

$\rightarrow$ Comparison between factual and counterfactual

**Fundamental problem of causal inference**: Analyst must infer counterfactual outcomes

## A Tale of Two Donors
\footnotesize

```{r, echo = FALSE}
charity %>% filter(user_id %in% c(13498, 13634))
```

\vspace{0.5cm}

Did donor 13498 not share about their donation **because** they received a standard solicitation?

## Notation

* **Unit** (indexed by $i$): individuals who have donated
* **Treatment variable** (causal variable of interest) $T_i$
    *  Received solicitation that emphasizes consequences of sharing for a cause
* **Treatment group** (treated units): Donors who recieve a "standard" solicitation
* **Control group** (untreated units): Donors who receive the "share for a cause" solicitation
* **Outcome variable(s)** (response variable) $Y_i$
    * `click-through`: Donor clicks on a pop-up encouraging sharing 
    * `recruitment`: Did donor's sharing lead to subsequent donation 

## Notation 

\begin{table}[]
\begin{tabular}{l|cc}
                                   & $T_i$ & $Y_i$ (click-through) \\
                                   \hline
Donor saw standard solicitation    & 0     & 0                     \\
Donor saw solicitation for a cause & 1     & 1                    \\
\hline
\end{tabular}
\end{table}

## Causal Effects and Counterfactuals 

* What does "$T_i$ causes $Y_i$ mean? $\rightsquigarrow$ **counterfactuals**, "*what if*"
    * Would donor $i$ who saw the standard solicitation have clicked through if they saw the one that emphasized the cause
* Two **potential outcomes**:
    * $Y_i(1)$: would donor click through if saw cause based solicitation?
    * $Y_i(0)$:  would donor click through if saw standard solicitation??
* **Causal effect**: $Y_i(1) - Y_i(0)$

* **Fundamental problem of causal inference**: only one of the two potential outcomes is observable per observation.

## Potential Outcomes 

\begin{table}[]
\begin{tabular}{l|cccc}
                                   & $T_i$ & $Y_i$ (click-through) & $Y_i(0)$ & $Y_i(1)$ \\
                                   \hline
Donor saw standard solicitation    & 0     & 0                     & 0        & ???      \\
Donor saw solicitation for a cause & 1     & 1                     & ???      & 1       \\
\hline
\end{tabular}
\end{table}

\vspace{0.5cm}

* **Association is not causation**
* How can we infer the missing counterfactuals?

## How to Figure Out Counterfactuals 

* **Need to find \alert{similar} observations**!

* Sounds easy ... but
    * Harder than it sounds, and 
    * Requires assumptions
    
* **Randomized experiments** are one possible solution**

## Summation Notation

* Define the **sample size** (number of observations) as $n$

* Therefore, we have $n$ measurements of some variable, $(Y_1, Y_2, \ldots, T_n)$

* We'll want to refer to the sum of these variables:

$$
Y_1 + Y_2 + Y_3 + Y_4 + \ldots + Y_{n-1} + Y_n
$$

* This is cumbersome to write down, so we'll use the *sigma* notation

$$
\sum_{i=1}^{n} Y_i = Y_1 + Y_2 + Y_3 + Y_4 + \ldots + Y_{n-1} + Y_n
$$
* $\sum_{i=1}^{n} Y_i$ says
    1. Initialize the running sum to the case when $i$ = 1.
    2. Increment $i$ by 1 and add the new expression to the running sum.
    3. Repeat step 2 until $i = n$.

## Averages 

* The **sample average** or **sample mean** is simply the sum of all values divided by the number of values

$$
\bar{Y} = \frac{1}{n} \sum_{i=1}^{n} Y_i
$$

* Suppose we surveyed six people, and 3 of them donated 20 dollars:

$$\bar{Y} = \frac{1}{6}(20 + 20 + 20 + 0 + 0 +0) = 10$$

## Quantity of Interest

We want to estimate the average causal effect over all units:

\alert{\textbf{Sample Average Treatment Effect}} (\alert{\textbf{SATE}}) $= \frac{1}{n}\sum_{i=1}^{n} [ Y_i(1) - Y_i(0)]$

What we can estimate instead:

\alert{\textbf{Difference in Means}} $= \bar{Y}_{\text{Treated}} - \bar{Y}_{\text{Control}}$

where: 

* $\bar{Y}_{\text{Treated}}$ is the observed average outcome in the treatment group
* $\bar{Y}_{\text{Control}}$ is the observed average outcome in the control group

* How do we ensure that the difference in means is a good estimate of the SATE?

## Randomized Control Trials 

**Randomization**!

Recall **Randomization** of the treatment makes the treatment and control groups "identical" on average.

* The two groups are similar in terms of all characteristics (both observed and unobserved).
    * Control group is similar to treatment group
    * $\rightsquigarrow$ outcome in control group $\approx$ what would have happened to treatment group if they had been in control group

## Potential Problems with RCTs

* **Placebo effects**:
    * Respondents will be affected by any intervention, even if they shouldn’t have any effect.
* **Hawthorne effects**:
    * Respondents act differently just knowing that they are under study.

## Did the Randomization *actually* Randomize? 

If it did, **we shouldn't see large differences** between treatment and control group on **pre-treatment variables**.

* Called "balance checking"

* Pre-treatment variable are those that are unaffected by treatment.
* We can check in the actual data for some pre-treatment variable $X$
    * $\bar{X}_{Treated}$: average value of variable for treated group.
    * $\bar{X}_{Control}$: average value of variable for control group.
* Under randomization, $\bar{X}_{Treated} - \bar{X}_{Control} \approx 0$

## Multiple Treatments 

* Instead of 1 treatment, we might have **multiple treatment arms**:
    * Control condition
    * Treatment A
    * Treatment B
    * Treatment C, etc
* In this case, we will look at multiple comparisons:
    * $\bar{Y}_{Treated, A} - \bar{Y}_{Control}$
    * $\bar{Y}_{Treated, B} - \bar{Y}_{Control}$
    * $\bar{Y}_{Treated, A} - \bar{Y}_{Treated, B}$

# Application: Encouraging Donors to Share About Charity

## Study Design: Setting

:::: {.columns}

::: {.column width="50%"}

\vspace{1.5cm}

**When**: Four-week period from August 13, 2020, to September 9, 2020

\vspace{0.5cm}

**Where**: DonorsChoose.org
:::

::: {.column width="50%"}
\centering
\includegraphics[width=6cm]{figs/BR-Donorschoose.jpg}

:::

::::

## Study Design: Treatments

**Control Condition**: 

> "Share this classroom with family and friends"

**Treatment Condition**:

> "Your donation can start a chain reaction, but only if you tell others about the cause. Share this classroom with family and friends"


## Study Design: Sharing Contributions

:::: {.columns}

::: {.column width="33%"}
\centering
\includegraphics[width=4cm]{figs/facebook.png}
:::

::: {.column width="33%"}
\centering
\includegraphics[width=4cm]{figs/twitter.png}
:::

::: {.column width="33%"}
\centering
\includegraphics[width=4cm]{figs/gmail.png}
:::

::::

## Visualizing the Outcomes

:::: {.columns}

::: {.column width="50%"}
```{r}
charity %>% 
    group_by(condition) %>%
    summarize(click = mean(clickthrough),
              std_err = sd(clickthrough) / sqrt(n())
              )%>%
    ggplot() +
    geom_bar(aes(x=condition, y=click), stat="identity", fill="skyblue", alpha=0.7) +
    geom_errorbar( aes(x=condition, ymin=click-std_err, ymax=click+std_err), width=0.4, colour="orange", alpha=0.9, size=1.5) +
    # scale_y_continuous(limits=c(.13,.155), oob = scales::squish) +
    theme_bw() + 
    ggtitle("Clickthrough Rate") +
    theme(text = element_text(size=28),
          plot.title = element_text(hjust = 0.5))

```
:::

::: {.column width="50%"}
```{r}
charity %>% 
    group_by(condition) %>%
    summarize(recruit = mean(recruited),
              std_err = sd(clickthrough) / sqrt(n())
              )%>%
    ggplot() +
    geom_bar(aes(x=condition, y=recruit), stat="identity", fill="skyblue", alpha=0.7) +
    geom_errorbar( aes(x=condition, ymin=recruit-std_err, ymax=recruit+std_err), width=0.4, colour="orange", alpha=0.9, size=1.5) +
    # scale_y_continuous(limits=c(.025,.035), oob = scales::squish) +
    theme_bw() +
    ggtitle("Recruitment Rate") +
    theme(text = element_text(size=28),
          plot.title = element_text(hjust = 0.5))

```
:::

::::

## Visualizing the Data ... and Cutting the axis

:::: {.columns}

::: {.column width="50%"}
```{r}
charity %>% 
    group_by(condition) %>%
    summarize(click = mean(clickthrough),
              std_err = sd(clickthrough) / sqrt(n())
              )%>%
    ggplot() +
    geom_bar(aes(x=condition, y=click), stat="identity", fill="skyblue", alpha=0.7) +
    geom_errorbar( aes(x=condition, ymin=click-std_err, ymax=click+std_err), width=0.4, colour="orange", alpha=0.9, size=1.5) +
    scale_y_continuous(limits=c(.13,.155), oob = scales::squish) +
    theme_bw() + 
    ggtitle("Clickthrough Rate") +
    theme(text = element_text(size=28),
          plot.title = element_text(hjust = 0.5))

```
:::

::: {.column width="50%"}
```{r}
charity %>% 
    group_by(condition) %>%
    summarize(recruit = mean(recruited),
              std_err = sd(clickthrough) / sqrt(n())
              )%>%
    ggplot() +
    geom_bar(aes(x=condition, y=recruit), stat="identity", fill="skyblue", alpha=0.7) +
    geom_errorbar( aes(x=condition, ymin=recruit-std_err, ymax=recruit+std_err), width=0.4, colour="orange", alpha=0.9, size=1.5) +
    scale_y_continuous(limits=c(.025,.035), oob = scales::squish) +
    theme_bw() +
    ggtitle("Recruitment Rate") +
    theme(text = element_text(size=28),
          plot.title = element_text(hjust = 0.5))

```
:::

::::

## Computing Proportions by Treatment


```{r, echo = TRUE}
charity %>% 
    group_by(condition) %>%
    summarize(click = mean(clickthrough),
              recruit = mean(recruited))
```

## Estimating SATEs: Clickthroughs

\begin{center}
\textbf{What's the \alert{correct} statistical test?}
\end{center}

```{r, echo = FALSE}
charity <-
    charity %>%
    mutate(clickthrough = as.logical(clickthrough),
           recruited = as.logical(recruited)
    )
```

* Test for `clickthroughs`
```{r, echo = TRUE}
prop_test(charity,
          clickthrough ~ condition
          )
```

## Estimating SATEs: Recruitment

\begin{center}
\textbf{What's the \alert{correct} statistical test?}
\end{center}


```{r, echo = TRUE}
prop_test(charity,
          recruited ~ condition
          )
```

## Regression as Mean Estimation

* Let $W_i$ be defined as follows:

$$
W_i =
\begin{cases}
1       \quad \text{if } i \text{ in the Treatment Group} \\
0       \quad \text{if } i \text{ in the Control Group} \\
\end{cases}
$$

Then we can write:

$$
Y_i = Y_i(0) + W_i (Y_i(1) - Y_i(0))
$$

## Regression as Mean Estimation

Recall the the definition of $\bar{Y}_{Control}$:

$$
\begin{aligned}
\bar{Y}_{Control} &= \frac{1}{n_0} \sum_{i=0}^{n_0} Y_i(0) \\
&= \hat{\beta_0}
\end{aligned}
$$

This is a consistent estimator of the expected value:

$$
\beta_0 = E[Y_i(0)]
$$


## Regression as Mean Estimation

Then, recall our definition of difference in means:

$$
\begin{aligned}
\text{Difference in Means} &= \bar{Y}_{\text{Treated}} - \bar{Y}_{\text{Control}} \\
& = \frac{1}{n_1} \sum_{i=1}^{n_1} Y_i(1) - \frac{1}{n_0} \sum_{i=1}^{n_0}Y_i(0) \\
& = \hat{\beta}_1
\end{aligned}
$$

which is a consistent estimator of the difference in means at the population level 

$$
\beta_1 = E[Y_i(1) - Y_i(0)]
$$

## Regression as Mean Estimation

Consider the regression equation:

$$
Y_i = \beta_0 + W_i \beta_1 + \varepsilon_i
$$

Then take expectations condition on treatment assignment:

$$
E(Y_i) = \beta_0 + W_i \beta_1
$$

This implies that we can estimate the ATE of a binary treatment via a linear regression of observed outcomes  $Y_i$ on a vector consisting of intercept and treatment assignment, $(1, W_i)$.

## Regression-based SATE

$$
clickthrough_i = \beta_0 + \beta_1 Condition_i + \varepsilon_i
$$

```{r, echo = TRUE}
clicks <- lm(clickthrough ~ condition, data = charity)
tidy(clicks)
```

**In class:** Interpret these coefficients.

## Regression-based SATE

$$
Recruit_i = \beta_0 + \beta_1 Condition_i + \varepsilon_i
$$

```{r, echo = TRUE}
recruit <- lm(recruited ~ condition, data = charity)
tidy(recruit)
```

**In class:** Interpret these coefficients.

##  A Balance Test?

The data provided by the authors **does not contain any information on pre-experiment variables.**

Thus, we **cannot test for balance** between control and treatment groups

# Wrap Up

## Summary 

* Causal inference with observational data is difficult due to omitted variable bias and selection effects
* Randomized Control Trials solve these issues and allow us to estimate a Sample Average Treatment Effect by randomly allocating units of observation to either an intervention or a control condition 
* SATE can be estimated from classical tools for statistical inference: t-tests, proportions tests and linear regression

## License & Citation
\small
Suggested Citation:

```{r, engine='out', echo=TRUE, eval = FALSE}
@misc{smwa2024_randomized,
      title={"Social Media and Web Analytics: Casuation and 
             Randomized Experiments"},
      author={Lachlan Deer},
      year={2024},
      url = "https://tisem-digital-marketing.github.io/2024-smwa"
}
```

This course adheres to the principles of the [\alertb{Open Science Community of Tilburg University}](https://www.tilburguniversity.edu/research/open-science-community). 
This initiative advocates for transparency and accessibility in research and teaching to all levels of society and thus creating more accountability and impact.

This work is licensed under a [\alertb{Creative Commons Attribution-ShareAlike 4.0 International License}](http://creativecommons.org/licenses/by-sa/4.0/).
